Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-07-24 chapters/Method.tex
Chapter two feedback Jens

\chapter{Method}
\label{chapter:method}
\epigraph{\textit{Du bist schön, aber dafür kannst du nichts.
Weder Lesen, noch Schreiben, noch was anderes}}{Alligatoah}
Based on the theory discussed in the previous chapter it is possible to build a model for numerical simulations.
Numerical simulations have been at the heart of fluid dynamics for a long time.
This is mainly because of two statements.
On the one hand we can not find an analytical solution to any given flow problem, that has to do with the shape and the structure of the Navier-Stokes equations.
In fact there are only a few flows that admit an analytical solution. 
They are often referred to as benchmark problems.
One of those is the famous ''Hagen–Poiseuille law'', see  Sec.~\ref{sec:theory_shallow_water}, which is an analytical solution for a flow field in a canal and was first derived by Poiseuille and Haagen in the 18th century~\cite{sutera1993history}.
There are a few other benchmark problems, especially in the low Reynolds number regime with a small or vanishing Stokes number.
The so called Stokes flows admit time reversal symmetry and many have theoretical solutions.
But going from laminar to turbulent flows with high Reynolds numbers poses still a problem, or as R. Feynman once said: ''... the most important unsolved problem of classical physics.''

@@

Coming up in this chapter is a short overview of numerical methods to simulate the thin film equation.
Starting with the straightforward finite differences approach and an introduction of the CFL criteria.
But also introducing the bottom up approach where Newtons equation of motion is solved for every molecule.
Followed by an short introduction to the main method of this work, the lattice Boltzmann method~\cite{PhysRevLett.56.1505, krugerLatticeBoltzmannMethod2017, succi}. 
The mathematical framework as well as it's link to kinetic theory will be discussed and the idea of the Chapmann-Enskog expansion will be outlined~\cite{Chapman, Enskog}. 
Based on the mathematical model it is possible to define a numerical approach.
Using assumptions, mainly concerning the collision operator, the resulting algorithm is fairly easy to implement and shows good scalability.
Which closes the loop with the above statement that the resulting source code is a perfect candidate for GPU programming.

\section{Numerical Methods}
\label{sec:numerical_methods}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{graphics/800px-Finite_Differences.svg.png}
    \caption{Discretization of a continuous function y (in red) along a discrete space x ($x_0, x_1, x_2$,...).
    Every point is shifted an interval distance h.
    %Such the discrete function y of $x\in\{x_0, x_1, x_2, ...\}$ is a straight connection between the dots.
    }
    \label{fig:finite_difference}
\end{figure}
The lattice Boltzmann method is one among many to simulate the behaviour of a fluid.
Especially when it comes to numerical simulations of the thin film equation, see Eq.~(\ref{eq:simple_thin_film}), other numerical approaches are more established and have been successfully applied for more then thirty years~\cite{becker2003complex, Peschka9275, PhysRevLett.95.244505, Mecke_2005, PhysRevE.63.011208, SCHWARTZ1998173}.
Since the thin film equation is a fourth order partial differential equation (PDE) one has to be careful in choosing an appropriate scheme.
Careful in the sense that not every numerical scheme will be able to provide first stability and second a correct approximation of the solution.
Problems for example appear when a simple Euler approach is used to solve the thin film equation.
The reason behind this complexity is the structure of the equation. 
It admits a fourth order differential in space, due to the pressure gradient, and allows for non linear behaviour.
Still though there are finite difference schemes that can be used to simulate the thin film equation.

@@

\subsection{Finite differences}
In fact there are further limitations to the problem of numerically approximating a partial differential equations.
Numerical methods for partial differential equations, e.g. the thin film or the wave equation, need to satisfy the Courant Friedrichs Lewy condition (CFL)~\cite{courant1928partiellen}.  
Which beyond other things states that choosing a increment for time can not be independent of the spatial resolution.
This can be written as
\begin{equation}\label{eq:CFL}
    C = \Delta t \left(\sum_{i=1}^n \frac{u_{x_i}}{\Delta x_i}\right) \leq C_{max},
\end{equation}
with $\Delta x_i$ and $\Delta t$ being the spatial and temporal increment, respectively and $u_{x_i}$ being the magnitude of the velocity in the respective direction.
Behind this concept is the idea that propagation of information has a upper limit.
For fluid dynamic problems one speaks about the speed of sound, in electrodynamics it would be the speed of light on the other hand.
Any propagation larger than this upper limit introduces stability issues, e.g. blow ups.
Thinking about a wave package, numerically it should be ensured that several time iterations happen before the package has reached the next point of the spatial resolved lattice.
$C_{max} = 1$ for an explicit solver to which for example the lattice Boltzmann method belongs (assuming a BGK collision operator).
But can be relaxed to be larger than one for implicit methods.
In Fig.~\ref{fig:finite_difference} there is a single lattice constant $h$ that sets the spatial resolution of the system as such $\Delta x = h$.
Time stepping thus the step from $u(x_i,t_0)$ to $u(x_i,t_1)$ can ensure that forces and therefore acceleration are well within the stability condition.
Concerning the thin film equation a rather well working scheme is the Crank-Nicolson approach~\cite{crank_nicolson_1947, PhysRevE.63.011208, 10.5555/1403886}.

@@

\subsection{Molecular dynamics}
Yet another approach that is the molecular dynamics method~\cite{haile1992molecular, PhysRevE.100.023108, doi:10.1063/1.1290698}.
Although when thinking of the motion molecules one does not think of the flow a \textbf{thin} film, molecular dynamics can be very insightful when dealing with thin film problems such as slip~\cite{jabbarzadeh1999wall, sega2013regularization}. 
Instead of working with macroscopic fields, such as density $\rho$ or velocity $\mathbf{u}$, the method uses a bottom up approach.
The main idea is to compute the trajectory of every particle using Newtons equation of motion, therefore assume particles are classical objects.
Interactions as well as boundary conditions can be added via potentials which act on particles.
A common interaction potential is the Lennard-Jones potential
\begin{equation}\label{eq:Len-Jon}
    U(r_{ij}) = 4\varepsilon\left[\left(\frac{\sigma}{r_{ij}}\right)^{12}-\left(\frac{\sigma}{r_{ij}}\right)^6\right],
\end{equation}
where $ij$ denotes the particle pair, $\varepsilon$ and $\sigma$ are the model dependent energy and length parameters respectively.
Clearly this approach is rather elegant.
Knowing the correct values for $\sigma$ and $\varepsilon$, the simulations show good agreement with experiments~\cite{PhysRevE.100.023108}.
However there is no free lunch. 
Because of the microscopic nature of this method it is compute intensive.
Per definition the method needs to know the position of every particle as well as its velocity.
But what really eats up all computing power is the computation of the potentials and therefore forces acting on the particles.
This surmounts to a lot of computational resources for small physical volumes, which is in the context of fluid flows and film formation problematic.
Assuming a box of a few nanometers in all directions and a physical time of about a picosecond ($10^{-12}s$) is already quite a lot to do for a common laptop.
Clearly a lot is depending on the implementation of the algorithm. 
There are well developed and supported software solutions that not only scale well but have an mature user interface, e.g. LAMMPS and GROMACS~\cite{PLIMPTON19951, BERENDSEN199543, lindahl2001gromacs}.  
However all development can not overcome that this method is suited only for relatively small systems and not for macroscopic scales.
On top comes the development of interaction potentials that due their complexity can not easily understood anymore.
Water for example although the most prominent liquid in our life is rather complex substance.
There exists not a single model for how water should interact but several, all with very fine tuned adjustments.
Without knowing why a certain water model is more suited than another one, this approach can come down to a black box.
Of course, similar to other parts of computational physics, all results need to be tested with an experiment but using molecular dynamics correct may require a lot more knowledge than e.g. finite differences.

@@

\subsection{Lattice Boltzmann and Chapman Enskog}
Since the method got first developed in the 1986 it has gained more and more popularity~\cite{PhysRevLett.56.1505}.
Due the simplicity of the method a lot has been worked out over last fourty years.
The reason why the lattice Boltzmann method is lacking behind when it comes to thin films flows is due to it's origin as an approximation to the Navier-Stokes equation.
That said there are of course studies of thin film flows using the lattice Boltzmann method, e.g. Sbragaglia et al.~\cite{Sbragaglia}.
However they rely on additional contributions for either multiphase or multicomponent models~\cite{shan1993lattice, shan_yuan_chen_2006, CHEN1995617}.
To actually solve Eq.~(\ref{eq:boltzmann_eq}) a few steps are still missing.
First being the discretization of the equation.
Now as the name suggests a lattice with lattice constants $\Delta x$ and $\Delta t$ for spatial and temporal discretization is the desired domain.
A first guess of this discretization on Eq.~(\ref{eq:boltzmann_eq}) could be
\begin{equation}\label{eq:LBM_discret_noforces}
    f_{\alpha}(\mathbf{x}+\mathbf{c}_{\alpha}\Delta t, t+ \Delta t) - f_{\alpha}(\mathbf{x}, t) = -\frac{\Delta t}{\tau}(f_{\alpha}(\mathbf{x}, t) - f^{eq}_{\alpha}(\mathbf{x}, t)),
\end{equation}
where the $\mathbf{c}_{\alpha}$ form a discrete set of $\alpha$ velocities.
Although this guess may seem different (\textit{simpler}) than Eq.~(\ref{eq:boltzmann_eq}) Chapmann and Enskog have shown that this does not only resembles the Boltzmann equation but also is an approximation to the Navier-Stokes equation~\cite{Chapman, Enskog}. 
The idea is to derive a perturbation expansion of the distribution functions around their equilibrium as such
\begin{equation}\label{eq:expansion_f}
    f_{\alpha} = f_{\alpha}^{eq} + \epsilon f_{\alpha}^{(1)} + \epsilon^2 f_{\alpha}^{(2)} + O(\epsilon^3), 
\end{equation}
where $\epsilon$ accounts for the order in Knudsen number ($Kn$). 
$Kn = \lambda/L$ is a measure to compare the mean free path to a characteristic length scale, i.e. ballistic and diffusive regime.
Introducing the non-equilibrium part of the distribution $f$ as $f^{neq} = f - f^{eq}$. 
They as well have to satisfy 
\begin{align}\label{eq:non_constraint}
    \sum_{\alpha} f_{\alpha}^{neq} = 0, \\
    \sum_{\alpha} \mathbf{c}_{\alpha}f_{\alpha}^{neq} = 0, 
\end{align}
where due to discretization the integration of Eqs.(\ref{eq:BGK_1}-\ref{eq:BGK_4}) becomes a sum.
The moments of the equilibrium functions on the other hand need to satisfy~\cite{chen1998lattice} 
\begin{align}
    \Pi^{eq} &= \sum_{\alpha} f_{\alpha}^{eq} = \rho, \label{eq:moments_equilibria_1}\\
    \Pi^{eq}_i &= \sum_{\alpha} c_{\alpha i}f_{\alpha}^{eq} = \rho u_i, \label{eq:moments_equilibria_2}\\
    \Pi^{eq}_{i j} &= \sum_{\alpha} c_{\alpha i} c_{\alpha j}f_{\alpha}^{eq} = \rho u_i u_j + \rho c_s^2\delta_{i j}, \label{eq:moments_equilibria_3}\\
    \Pi^{eq}_i &= \sum_{\alpha} c_{\alpha i} c_{\alpha j} c_{\alpha k}f_{\alpha}^{eq} = \rho c_s^2(u_i\delta_{j k} + u_j\delta_{k i} + u_k\delta_{i j}), \label{eq:moments_equilibria_4}
\end{align}
In fact for every order of expansion Eq.~(\ref{eq:expansion_f}) one gets
\begin{align}\label{eq:all_order_constraint}
    \sum_{\alpha} f_{\alpha}^{(n)} = 0, \\
    \sum_{\alpha} \mathbf{c}_{\alpha}f_{\alpha}^{(n)} = 0. 
\end{align}

@@

The idea however of this thesis is not to approximate the Navier-Stokes equation but have a numerical solver for the thin film equation.
Luckily for me and as outlined in Chap.~\ref{chapter:theory} there is a system which has some similarities with the thin film equation.
That system is the shallow water equation. 
In fact it is possible to derive with an similar ansatz, e.g. a slow mode expansion, the shallow water equation from the Boltzmann equation~\cite{Salmon:1999:0022-2402:503, zhou2004lattice, PhysRevE.65.036309}.
While Eq.~(\ref{eq:LBM_discret_noforces}) does not need to be modified the shallow water equations require a different equilibrium distribution.
Instead of an isothermal equation of state $p = \rho c_s^2$, the shallow water system uses the hydrostatic pressure Eq.~(\ref{eq:hydro_static_int}) as closure.

>>>>

\chapter{Method}
\label{chapter:method}
\epigraph{\textit{Du bist schön, aber dafür kannst du nichts.
Weder Lesen, noch Schreiben, noch was anderes}}{Alligatoah}
Based on the theory discussed in the previous chapter it is possible to build a model for numerical simulations.
Numerical simulations have been at the heart of fluid dynamics for a long time.
This is mainly because of two statements.
On the one hand we can not find an analytical solution to any given flow problem, that has to do with the shape and the structure of the Navier-Stokes equations.
In fact there are only a few flows that admit an analytical solution. 
They are often referred to as benchmark problems.
One of those is the famous ''Hagen–Poiseuille law'', see  Sec.~\ref{sec:theory_shallow_water}, which is an analytical solution for a flow field in a canal and was first derived by Poiseuille and Haagen in the 18th century~\cite{suteraHistoryPoiseuilleLaw}.
There are a few other benchmark problems, especially in the low Reynolds number regime with a small or vanishing Stokes number.
The so called Stokes flows admit time reversal symmetry and many have theoretical solutions.
But going from laminar to turbulent flows with high Reynolds numbers poses still a problem, or as R. Feynman once said: ''... the most important unsolved problem of classical physics.''

@@

Coming up in this chapter is a short overview of numerical methods to simulate the thin film equation.
Starting with the straightforward finite differences approach and an introduction of the CFL criteria.
But also introducing the bottom up approach where Newtons equation of motion is solved for every molecule.
Followed by an short introduction to the main method of this work, the lattice Boltzmann method~\cite{PhysRevLett.56.1505, krugerLatticeBoltzmannMethod2017, succi}. 
The mathematical framework as well as it's link to kinetic theory will be discussed and the idea of the Chapman-Enskog expansion will be outlined~\cite{chapmanMathematicalTheoryNonuniform1990, enskogKinetischeTheorieVorgange1917}. 
Based on the mathematical model it is possible to define a numerical approach.
Using assumptions, mainly concerning the collision operator, the resulting algorithm is fairly easy to implement and shows good scalability.
Which closes the loop with the above statement that the resulting source code is a perfect candidate for GPU programming.

\section{Numerical Methods}
\label{sec:numerical_methods}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{graphics/800px-Finite_Differences.svg.png}
    \caption{Discretization of a continuous function y (in red) along a discrete space x ($x_0, x_1, x_2$,...).
    Every point is shifted an interval distance h.
    %Such the discrete function y of $x\in\{x_0, x_1, x_2, ...\}$ is a straight connection between the dots.
    }
    \label{fig:finite_difference}
\end{figure}
The lattice Boltzmann method is one among many to simulate the behaviour of a fluid.
Especially when it comes to numerical simulations of the thin film equation, see Eq.~(\ref{eq:simple_thin_film}), other numerical approaches are more established and have been successfully applied for more then thirty years~\cite{beckerComplexDewettingScenarios2003, peschkaSignaturesSlipDewetting2019, davidovitchSpreadingViscousFluid2005, meckeThermalFluctuationsThin2005, diezGlobalModelsMoving2000, schwartzSimulationDropletMotion1998}.
Since the thin film equation is a fourth order partial differential equation (PDE) one has to be careful in choosing an appropriate scheme.
Careful in the sense that not every numerical scheme will be able to provide first stability and second a correct approximation of the solution.
Problems for example appear when a simple Euler approach is used to solve the thin film equation.
The reason behind this complexity is the structure of the equation. 
It admits a fourth order differential in space, due to the pressure gradient, and allows for non linear behaviour.
Still though there are finite difference schemes that can be used to simulate the thin film equation.

@@

\subsection{Finite differences}
In fact there are further limitations to the problem of numerically approximating a partial differential equations.
Numerical methods for partial differential equations, e.g. the thin film or the wave equation, need to satisfy the Courant Friedrichs Lewy condition (CFL)~\cite{courant1928partiellen}.  
Which beyond other things states that choosing a increment for time can not be independent of the spatial resolution.
This can be written as
\begin{equation}\label{eq:CFL}
    C = \Delta t \left(\sum_{i=1}^n \frac{u_{x_i}}{\Delta x_i}\right) \leq C_{max},
\end{equation}
with $\Delta x_i$ and $\Delta t$ being the spatial and temporal increment, respectively and $u_{x_i}$ being the magnitude of the velocity in the respective direction.
Behind this concept is the idea that propagation of information has a upper limit.
For fluid dynamic problems one speaks about the speed of sound, in electrodynamics it would be the speed of light on the other hand.
Any propagation larger than this upper limit introduces stability issues, e.g. blow ups.
Thinking about a wave package, numerically it should be ensured that several time iterations happen before the package has reached the next point of the spatial resolved lattice.
$C_{max} = 1$ for an explicit solver to which for example the lattice Boltzmann method belongs (assuming a BGK collision operator).
But can be relaxed to be larger than one for implicit methods.
In Fig.~\ref{fig:finite_difference} there is a single lattice constant $h$ that sets the spatial resolution of the system as such $\Delta x = h$.
Time stepping thus the step from $u(x_i,t_0)$ to $u(x_i,t_1)$ can ensure that forces and therefore acceleration are well within the stability condition.
Concerning the thin film equation a rather well working scheme is the Crank-Nicolson approach~\cite{crank_nicolson_1947, diezGlobalModelsMoving2000, 10.5555/1403886}.

@@

\subsection{Molecular dynamics}
Yet another approach that is the molecular dynamics method~\cite{haile1992molecular, zhangMolecularSimulationThin2019, doi:10.1063/1.1290698}.
Although when thinking of the motion molecules one does not think of the flow a \textbf{thin} film, molecular dynamics can be very insightful when dealing with thin film problems such as slip~\cite{jabbarzadeh1999wall, sega2013regularization}. 
Instead of working with macroscopic fields, such as density $\rho$ or velocity $\mathbf{u}$, the method uses a bottom up approach.
The main idea is to compute the trajectory of every particle using Newtons equation of motion, therefore assume particles are classical objects.
Interactions as well as boundary conditions can be added via potentials which act on particles.
A common interaction potential is the Lennard-Jones potential
\begin{equation}\label{eq:Len-Jon}
    U(r_{ij}) = 4\varepsilon\left[\left(\frac{\sigma}{r_{ij}}\right)^{12}-\left(\frac{\sigma}{r_{ij}}\right)^6\right],
\end{equation}
where $ij$ denotes the particle pair, $\varepsilon$ and $\sigma$ are the model dependent energy and length parameters respectively.
Clearly this approach is rather elegant.
Knowing the correct values for $\sigma$ and $\varepsilon$, the simulations show good agreement with experiments~\cite{zhangMolecularSimulationThin2019}.
However there is no free lunch. 
Because of the microscopic nature of this method it is compute intensive.
Per definition the method needs to know the position of every particle as well as its velocity.
But what really eats up all computing power is the computation of the potentials and therefore forces acting on the particles.
This surmounts to a lot of computational resources for small physical volumes, which is in the context of fluid flows and film formation problematic.
Assuming a box of a few nanometers in all directions and a physical time of about a picosecond ($10^{-12}s$) is already quite a lot to do for a common laptop.
Clearly a lot is depending on the implementation of the algorithm. 
There are well developed and supported software solutions that not only scale well but have an mature user interface, e.g. LAMMPS and GROMACS~\cite{PLIMPTON19951, BERENDSEN199543, lindahl2001gromacs}.  
However all development can not overcome that this method is suited only for relatively small systems and not for macroscopic scales.
On top comes the development of interaction potentials that due their complexity can not easily understood anymore.
Water for example although the most prominent liquid in our life is rather complex substance.
There exists not a single model for how water should interact but several, all with very fine tuned adjustments.
Without knowing why a certain water model is more suited than another one, this approach can come down to a black box.
Of course, similar to other parts of computational physics, all results need to be tested with an experiment but using molecular dynamics correct may require a lot more knowledge than e.g. finite differences.

@@

\subsection{Lattice Boltzmann and Chapman Enskog}
Since the method got first developed in the 1986 it has gained more and more popularity~\cite{PhysRevLett.56.1505}.
Due the simplicity of the method a lot has been worked out over last fourty years.
The reason why the lattice Boltzmann method is lacking behind when it comes to thin films flows is due to it's origin as an approximation to the Navier-Stokes equation.
That said there are of course studies of thin film flows using the lattice Boltzmann method, e.g. Sbragaglia et al.~\cite{Sbragaglia}.
However they rely on additional contributions for either multiphase or multicomponent models~\cite{shan1993lattice, shan_yuan_chen_2006, CHEN1995617}.
To actually solve Eq.~(\ref{eq:boltzmann_eq}) a few steps are still missing.
First being the discretization of the equation.
Now as the name suggests a lattice with lattice constants $\Delta x$ and $\Delta t$ for spatial and temporal discretization is the desired domain.
A first guess of this discretization on Eq.~(\ref{eq:boltzmann_eq}) could be
\begin{equation}\label{eq:LBM_discret_noforces}
    f_{\alpha}(\mathbf{x}+\mathbf{c}_{\alpha}\Delta t, t+ \Delta t) - f_{\alpha}(\mathbf{x}, t) = -\frac{\Delta t}{\tau}(f_{\alpha}(\mathbf{x}, t) - f^{eq}_{\alpha}(\mathbf{x}, t)),
\end{equation}
where the $\mathbf{c}_{\alpha}$ form a discrete set of $\alpha$ velocities.
Although this guess may seem different (\textit{simpler}) than Eq.~(\ref{eq:boltzmann_eq}) Chapman and Enskog have shown that this does not only resembles the Boltzmann equation but also is an approximation to the Navier-Stokes equation~\cite{chapmanMathematicalTheoryNonuniform1990, enskogKinetischeTheorieVorgange1917}. 
The idea is to derive a perturbation expansion of the distribution functions around their equilibrium as such
\begin{equation}\label{eq:expansion_f}
    f_{\alpha} = f_{\alpha}^{eq} + \epsilon f_{\alpha}^{(1)} + \epsilon^2 f_{\alpha}^{(2)} + O(\epsilon^3), 
\end{equation}
where $\epsilon$ accounts for the order in Knudsen number ($Kn$). 
$Kn = \lambda/L$ is a measure to compare the mean free path to a characteristic length scale, i.e. ballistic and diffusive regime.
Introducing the non-equilibrium part of the distribution $f$ as $f^{neq} = f - f^{eq}$. 
They as well have to satisfy 
\begin{align}\label{eq:non_constraint}
    \sum_{\alpha} f_{\alpha}^{neq} = 0, \\
    \sum_{\alpha} \mathbf{c}_{\alpha}f_{\alpha}^{neq} = 0, 
\end{align}
where due to discretization the integration of Eqs.(\ref{eq:BGK_1}-\ref{eq:BGK_4}) becomes a sum.
The moments of the equilibrium functions on the other hand need to satisfy~\cite{chen1998lattice} 
\begin{align}
    \Pi^{eq} &= \sum_{\alpha} f_{\alpha}^{eq} = \rho, \label{eq:moments_equilibria_1}\\
    \Pi^{eq}_i &= \sum_{\alpha} c_{\alpha i}f_{\alpha}^{eq} = \rho u_i, \label{eq:moments_equilibria_2}\\
    \Pi^{eq}_{i j} &= \sum_{\alpha} c_{\alpha i} c_{\alpha j}f_{\alpha}^{eq} = \rho u_i u_j + \rho c_s^2\delta_{i j}, \label{eq:moments_equilibria_3}\\
    \Pi^{eq}_i &= \sum_{\alpha} c_{\alpha i} c_{\alpha j} c_{\alpha k}f_{\alpha}^{eq} = \rho c_s^2(u_i\delta_{j k} + u_j\delta_{k i} + u_k\delta_{i j}), \label{eq:moments_equilibria_4}
\end{align}
In fact for every order of expansion Eq.~(\ref{eq:expansion_f}) one gets
\begin{align}\label{eq:all_order_constraint}
    \sum_{\alpha} f_{\alpha}^{(n)} = 0, \\
    \sum_{\alpha} \mathbf{c}_{\alpha}f_{\alpha}^{(n)} = 0. 
\end{align}

@@

The idea however of this thesis is not to approximate the Navier-Stokes equation but have a numerical solver for the thin film equation.
Luckily for me and as outlined in Chap.~\ref{chapter:theory} there is a system which has some similarities with the thin film equation.
That system is the shallow water equation. 
In fact it is possible to derive with an similar ansatz, e.g. a slow mode expansion, the shallow water equation from the Boltzmann equation~\cite{salmonLatticeBoltzmannMethod1999, zhouLatticeBoltzmannMethods2004, dellarNonhydrodynamicModesPriori2002}.
While Eq.~(\ref{eq:LBM_discret_noforces}) does not need to be modified the shallow water equations require a different equilibrium distribution.
Instead of an isothermal equation of state $p = \rho c_s^2$, the shallow water system uses the hydrostatic pressure Eq.~(\ref{eq:hydro_static_int}) as closure.
